Characterizing Stellar Streams by ML

### 2023-08-30 src/tex/ms.tex
[showyourwork] overleaf sync

\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}
\usepackage{amsmath, amssymb}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{dsfont}
\usepackage[inline]{enumitem}
\usepackage{cleveref}
\usepackage{todonotes}

@@

    context. \citet{2022MNRAS.514.1757P} developed a data-driven spline-based

@@

    \citet{2019AJ....158..223P} utilized a mixture model to fit the globular

@@

        A stellar stream is the association of stars tidally stripped from a
        common progenitor system that orbits in an underlying galactic
        gravitational field \citep{Johnston1998, HelmiWhite1999}. Stream
        progenitors are gravitationally bound systems, such as satellite dwarf
        galaxies or star clusters \citep{Odenkirchen2001, Majewski2003}. In the
        absence of external interactions these systems are stable, though still
        capable of mass loss driven by internal processes \citep{Hills1975,
        HeggieHut2003}, e.g. relaxation and binary interactions. Considering
        external interactions, these progenitor systems may be disrupted:
        completely, as in the merger events that formed the Galactic stellar
        halo \citep{Lynden-Bell1967, Searle1978}; or partially, such as tidal
        disruption forming a stellar stream.

        In the approximation of a smooth galactic gravitational field, the
        progenitor's extent is limited to the tidal radius. Through processes
        internal and external, progenitor stars can achieve sufficient energy to
        bring them beyond the tidal boundary, thus ``escaping'' into the
        galactic field. Escape primarily occurs at both an inner and outer
        saddle point of the combined potential -- e.g. the L2 and L3 Lagrange
        points for a spherical galactic field \citep{Ross1997, Fukushige2000,
        BinneyTremaine2008}. Generally, the initial phase space position of the
        escaped star is very similar to that of the progenitor: the
        configuration position is the tidal boundary and the star has a low
        peculiar velocity. Since both the escaped star and its progenitor orbit
        in the same underlying galactic potential and have similar phase space
        positions their orbits are similar \citep{BinneyTremaine2008}. Given
        sufficient time the phase space separation will increase and an escaped
        star will lead or trail the position of the progenitor, depending on the
        location from which it escaped. Therefore, a collection of tidally
        stripped stars will form extended structures, aka tails or streams, that
        are few to thousands of parsecs in length \citep{Johnston1998,
        HelmiWhite1999, Bovy2014}.

        To be a coherent, observable structure the peculiar velocity dispersion
        of escaped stars relative to the progenitor must be relatively low, but
        does vary from system to system. Stars in kinematically hot streams
        (e.g. from progenitors with larger kinematic dispersions) have more
        dissimilar orbits from each other and the progenitor than do stars in
        cold streams, with smaller dispersions \citep{Johnston1998, Hendel2015}.
        The orbit of a star depends on its phase space position and the
        gravitational potential; therefore measuring a star's position and
        velocity and observing its orbit reveals the local potential, which is
        normally dominated by the Galaxy (primarily the dark matter halo at
        large distances) \citep{Gibbons2014}. However, an orbit is impractical
        to observe on human timescales. Cold streams offer a workaround.  Like
        measuring one star at many points in one orbit over an extended period,
        streams are many stars measured at different points along approximately
        one orbit at the same time. Therefore stellar streams are one of the
        most promising means to study the Galactic gravitational potential: the
        stellar components as well as the dark matter \citep{Johnston2016}.

        On the observational front there has been a large concerted effort in
        the field to find new stellar streams \citep[for an extensive list of
        known Galactic streams see][]{Mateu2022}, extend the detected extents of
        known streams \citep[e.g. for \stream{Palomar 5} alone:][]{Rockosi2002,
        Ibata2017, Grillmair2006, Carlberg2012, StarkmanEtAl2019}, and increase
        the general purity of stream catalogs.

@@

        \begin{enumerate}
            \item Streamfinder \citep{STREAMFINDER}. This method requires the potential
            \item VIA MACHINAE \citep{2022MNRAS.509.5992S}
            \item Kiyan and Adrian
            \item Matched Filter
            \item Gaia stuff: Adrian and Ana and Sara et al
            \item Uniform Modeling of 12 Stellar stream \citep{PatrickEtAl2022}. Only on phi1, phi2, and CMD. Their CMD method is more sophisticated. Worth discussing.
        \end{enumerate}

@@

        However, data observations are not taken from the GCOM, but from  $\sim
        8$ kpc distant \citep{GRAVITY2018, Leung2022} in a local frame,
        typically from vantage points such as Earth or its solar Lagrange
        points, and subsequently represented in the International Celestial
        Reference System (ICRS) \citep{ICRS1997}.  Restricted as we are to a
        very local volume, there is no means to observe systems from the more
        ``intrinsic" reference frames. In our observational frames there is no
        inherent assurance of effecting a rotation that would similarly align
        the stream within observer coordinates such that $\phi_2(\phi_1) \approx
        0$.  Thankfully, most streams are adequately distant from both the
        Galactic center and the observer, or exhibit orientations such that a
        rotation a) can be established or b) holds valid. For most of the
        best-studied globular cluster streams -- \stream{GD-1}, \stream{Palomar
        5}, \stream{Jhelum}, etc. -- this holds true. However for some streams
        of dwarf galaxies, like the Magellanic stream which wraps multiple times
        around the Milky Way \citep{Wannier1972}, no such transformation is
        possible. More generally, a stream's ``sky"-projected path may be
        kinked, e.g. by subhalo interactions, such that it is a many-to-one
        function in $\phi_1$.

        In this work we treat with the vast majority of streams, which have a
        ``sky" transformation such that the stream is an injective function in
        $\phi_2(\phi_1)$.  Since we are examining known streams this
        transformation is known \textit{a priori}, or may be constructed during
        pre-processing \citep[see][\S2.1]{StarkmanEtAl2023}.  We note that, with
        sufficient prior knowledge, many-to-one functions may be broken into
        injective segments and their connection constrained by a prior. Thus the
        methods developed subsequently may extented and applied even to wrapped
        and kinked streams. However, we leave extending the model framework to a
        future work.

@@

        Mixture models (MMs) are a statistical method to represent a population
        as a set of a sub-populations specified by probability distributions.
        These distributions are characterized by their parameters and the
        mixture by a corresponding set of mixture coefficients. Scalar mixture
        models may be extended to model completely general conditional
        distributions by making the distribution parameters and mixture
        coefficients into functions of a conditional input
        \citep{McLachlan1989MixtureM}.  Mixture Density Networks (MDNs) are a
        machine learning method to implement these general MM, taking the
        parameters and coefficients to be the outputs of feed-forward neural
        networks and as such, general continuous functions \citep{Bishop1994}.
        Functional parameters allow the model to capture variation over those
        parameters that is not possible with scalar-valued models. Moreover, by
        using neural networks the inferred functional forms are ``model-free"
        and driven entirely by the data.

@@

        Streams have a diversity of measurement coverage: several have full 6-D
        astrometrics as well as photometrics and chemical abundances available
        \citep[e.g.,][]{Koposov2019, Antoja2020, 2022ApJ...928...30L},
        while most have partial coverage, lacking some feature dimensions.
        % TODO: terminology ND-phase-space vs feature columns
        Most streams in \Gaia {\color{red} [cite]}, for instance, have proper
        motions but not radial velocities nor reliable parallax measurements.
        Followup surveys, like \dataarchive{S5} {\color{red} [cite]}, add many
        feature dimensions, however only for a select set of streams. Given the
        diversity of coverage it is important that the likelihood model --
        \autoref{eq:membership_prob} -- account for missing data.

@@

            % link to the code!
            In the code we offer a few ansatz for the PDMF, though any
            user-defined function may be used instead. The default assumption,
            known to be non-physical, is a uniform distribution over $\gamma$.
            Similarly non-physical are truncated uniform distributions,
            considering only a segment of the isochrone, e.g. the main sequence
            turnoff (MSTO). We include also a Kroupa initial mass function (IMF)
            \citep{2001MNRAS.322..231K}. The Kroupa IMF, and all
            physically-motivated IMFs, have an much larger fraction of low-mass
            stars then high-mass stars, the latter being less likely to form.
            This is observed in streams, particularly their progenitors. However
            the Kroupa IMF, and all IMF, are poor fits to the observed PDMF as
            stream progenitors like globular clusters are extremely old
            ($\sim10$ Gyr) while an IMF is the zero-age distribution.  How then
            should the PDMF $\prior_{\cal{I}}$ be treated?
            % ⬆ rephrase previous sentence
            One approach to the PDMF is to use a more informed distribution, for
            example time-evolving an IMF, either analytically or by simulation,
            to the age of the stream's stellar population.  As discussed
            previously, globular clusters are known to be poorly fit by
            theoretic isochrones with realistic inputs (e.g. age, $Z$). Thus
            fitting an isochrone's track to the data means its age, metallicity
            and therefore PDMF are not well fit \TODO{[cite]}. Moreover, for
            many streams, e.g. those lacking known progenitors, the PDMF is
            observationally poorly constrained.  In short, using any fixed
            distribution is simple in implementation, however the choice of
            distribution is not.

@@

        We generate synthetic stream observations by sampling uniformly in $\phi_1$
        In $\phi_2$ we sample from a quadratic in $\phi_1$ with normally-distributed
        $0.15^{\circ}$ scatter.  The distances are sampled from a linear funciton in
        $\phi_1$ with normally-distributed $0.25 \ \unit{kpc}$ scatter, then
        transformed to the parallax $\parallax$.  Non-poissonian density variations
        are then simulated by evaluating the stream density with two Gaussian
        distributions centered at different values of $\phi_1$ with random
        amplitudes (which sum to $1$) and widths.  Stream stars are then removed
        using random sampling, weighted by their likelihood of belonging to the
        density. Post removal, the number of stream stars is
        $\variable{output/mock/nstream_variable.txt}$.
        $\variable{output/mock/nbackground_variable.txt}$ background astrometric points
        are generated from exponential distributions at each $\phi_1$.

        The synthetic stream's photometry is also simulated using a
        \variable{output/mock/isochrone_age_variable.txt},
        [Fe/H] = \variable{output/mock/isochrone_feh_variable.txt} MIST isochrone
        \TODO{[cite brutus]}. The isochrone is truncated just
        short of the giant branch, and is sampled from uniformly along the isochrone
        (i.e., the initial mass function, $P(\gamma)$, is uniform) as well as with
        an intrinsic uniform 0.2~dex scatter orthogonal to the isochrone track. The
        isochrone is shifted by the stream's parallax modulus, which is derived from
        the astrometric distance track.  The background photometric coordinates are
        generated as a 2-dimensional Gaussian with non-zero covariance.  Thus, the
        synthetic data scenario described here generates a mock stream in
        astrometric coordinates and photometric coordinates, with density
        variations, a distance gradient which is consistent in parallax and distance
        modulus (i.e., magnitudes), all superimposed over a noisy field of
        background points in each astrometric and photometric dimension.

@@

        We select on-stream and off-stream regions using the known $\phi_1, \phi_2$
        range of the mock stream.  The upper panel of
        \autoref{fig:mock_data_photometric_background_selection} shows these
        selections blue and black, respectively.  The lower panel shows the on and
        off-stream selections in photometric coordinates.  In this space it is
        apparent that the on-stream region has two populations, drawn from the
        isochrone and from background points included in the simple $\phi_1, \phi_2$
        box.  In contrast the off-stream region, has only the the background population.
        We use the off-stream population to train a conditional normalizing flow
        (see \autoref{sub:method:photometric_model:off_stream}) on the magnitudes
        $g$ and $r$ and conditioned on $\phi_1$.  The normalizing flow is now a
        non-parametric distribution characterizing the photometric background,
        including its variation over $\phi_1$.  For this mock stream the flexibility
        of the normalizing flow greatly exceeds necessity: the background is
        drawn from a $\phi_1$-invariant multivariate normal, essentially the base
        distribution of the normalizing flow.  It is therefore no surprise that
        the trained normalizing flow provides an excellent characterization of
        the background.  We will later see that the flow performs similarly well
        on complex distributions.  Having trained the background photometric model
        we proceed to build and train the full model.

        Per \autoref{eq:stream_and_bkg_prob} the stream is a conditionally independent
        addition of both astrometric and photometric models.
        In astrometrics the stream is a Gaussian track in $\phi_2$ and
        parallax $\varpi$, using the model defined in
        \autoref{ssub:method:astrometric_model:on_stream}. We use a 4-layer 
        \texttt{LinTanh} with shape (1 input, 64 hidden, 4 output), where the 4 outputs
        are $[\mu_{\phi_2}, \sigma_{\phi_2}, \mu_{\varpi}, \sigma_{\varpi}](\phi_1)$.
        The network is small, but sufficiently large to permit 15\% dropout and retain
        robust predictions.
        The photometric model is a distance modulus-shifted isochrone, using the model
        outlined in  \autoref{ssub:method:photometric_model:on_stream}. We use the
        same isochrone parameters as were used to simulate the stream. For this mock
        stream we demonstrate simultaneous astrometric and photometric modeling, linking
        the distance modulus and parallax track, as explained in
        \autoref{ssub:method:linking_to_astrometrics}. Thus the photometric model
        does not have its own network, using instead the outputs
        $[\mu_{\varpi}, \sigma_{\varpi}](\phi_1)$.
        The background model is split in two components, the first is a uniform distribution
        in $\phi_2$, the second is an exponential distribution for $\varpi$.
        The uniform distribution does not have any parameters, and thus no neural network.
        The exponential distribution has only the slope parameter and we use a 3-layer 
        \texttt{LinTanh} with shape (1, 32, 1) and 15\% dropout. Background and
        stream are combined in a mixture model whose mixture parameters are a 4-layer 
        (1, 64, 1) \texttt{LinTanh} network. We train the mixture model with an
        \texttt{AdamW} optimizer with learning rate of $5\times10^{-3}$ and a scheduler
        to periodically and temporarily increase the learning rate, encouraging the optimizer
        towards better minima.

@@

        % \begin{enumerate}
        %     \item Gaia data.
        %         \begin{enumerate}
        %             \item box size
        %             \item selection criteria
        %         \end{enumerate}
        %     \item crossmatched with panstarrs1
        %     \item extinction corrected with bayestar2019
        % \end{enumerate}

        We utilize data from \Gaia{} DR3 for the astrometric coordinates of
        each star, and \PanStarrs{} to obtain more accurate color information.

        We cross-match the \Gaia{} field on \PanStarrs{} using the provided
        \texttt{Best-neighbors} catalog.

        adopt broad cuts on the cross-matched data to limit the field to a
        region known to contain GD-1. We perform these rough cuts by requiring...

        We use the \texttt{GD1Koposov10} frame defined in Gala \TODO{[cite gala and Koposov10]}

        \JN{fill in what the cuts are and reference proper motion figure and CMD
        figure.}

        The result is a field populated by \JN{x} stars, from which we would
        like to identify stream members.

        \todo[inline]{Dust}
            We use \package{dustmaps} Bayestar 19 for the dust modeling. We treat it as known. Include as a source of error in $\Sigma_n$
            % TODO: actually implement that in the model

@@

        We model GD-1 as a three-component mixture consisting of a main stream
        component, a spur, and background. In order to guide our model towards
        GD-1, we set down 4 ``control points" along GD-1's main track, and 3
        control points along the spur (illustrated in \JN{x}). These control
        points act as Gaussian priors on the model. We set the Gaussian error on
        these control points to large values, so that their influence on the
        model is not dominant. Indeed, we find that the value of the control
        points is to help reduce training time, since otherwise it will take the
        flexible density model longer time to converge to the actual stream.

        The weight parameter is also set to truncate to $0$ at $\phi_1 < X$ and $\phi_1 > Y$. This ensures that we do not model stream stars that fall

@@

        \begin{figure*}[h]
            \script{gd1/plot/data_selection.py}
            \centering
            \includegraphics[width=1\linewidth]{figures/gd1/data_selection.pdf}
            \caption{Caption}
            \label{fig:gd1-data_selection}
        \end{figure*}

@@

        \subsubsection{VIA MACHINAE}{\cite{2022MNRAS.509.5992S}}

>>>>

\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}
\usepackage{amsmath, amssymb}
\usepackage{gensymb}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{dsfont}
\usepackage[inline]{enumitem}
\usepackage{cleveref}
\usepackage{todonotes}

@@

    context. \citet{Patrick+2022} developed a data-driven spline-based

@@

    \citet{Price-Whelan+2019} utilized a mixture model to fit the globular

@@

        A stellar stream is the association of stars tidally stripped from a
        common progenitor system that orbits in an underlying galactic
        gravitational field \citep{Johnston1998, HelmiWhite1999}. Stream
        progenitors are gravitationally bound systems, such as satellite dwarf
        galaxies or star clusters \citep{Odenkirchen+2001, Majewski+2003}. In the
        absence of external interactions these systems are stable, though still
        capable of mass loss driven by internal processes \citep{Hills1975,
        HeggieHut2003}, e.g. relaxation and binary interactions. Considering
        external interactions, these progenitor systems may be disrupted:
        completely, as in the merger events that formed the Galactic stellar
        halo \citep{Lynden-Bell1967, SearleZinn1978}; or partially, such as tidal
        disruption forming a stellar stream.

        In the approximation of a smooth galactic gravitational field, the
        progenitor's extent is limited to the tidal radius. Through processes
        internal and external, progenitor stars can achieve sufficient energy to
        bring them beyond the tidal boundary, thus ``escaping'' into the
        galactic field. Escape primarily occurs at both an inner and outer
        saddle point of the combined potential -- e.g. the L2 and L3 Lagrange
        points for a spherical galactic field \citep{Ross+1997, FukushigeHeggie2000,
        BinneyTremaine2008}. Generally, the initial phase space position of the
        escaped star is very similar to that of the progenitor: the
        configuration position is the tidal boundary and the star has a low
        peculiar velocity. Since both the escaped star and its progenitor orbit
        in the same underlying galactic potential and have similar phase space
        positions their orbits are similar \citep{BinneyTremaine2008}. Given
        sufficient time the phase space separation will increase and an escaped
        star will lead or trail the position of the progenitor, depending on the
        location from which it escaped. Therefore, a collection of tidally
        stripped stars will form extended structures, aka tails or streams, that
        are few to thousands of parsecs in length \citep{Johnston1998,
        HelmiWhite1999, Bovy2014}.

        To be a coherent, observable structure the peculiar velocity dispersion
        of escaped stars relative to the progenitor must be relatively low, but
        does vary from system to system. Stars in kinematically hot streams
        (e.g. from progenitors with larger kinematic dispersions) have more
        dissimilar orbits from each other and the progenitor than do stars in
        cold streams, with smaller dispersions \citep{Johnston1998, HendelJohnston2015}.
        The orbit of a star depends on its phase space position and the
        gravitational potential; therefore measuring a star's position and
        velocity and observing its orbit reveals the local potential, which is
        normally dominated by the Galaxy (primarily the dark matter halo at
        large distances) \citep{Gibbons+2014}. However, an orbit is impractical
        to observe on human timescales. Cold streams offer a workaround.  Like
        measuring one star at many points in one orbit over an extended period,
        streams are many stars measured at different points along approximately
        one orbit at the same time. Therefore stellar streams are one of the
        most promising means to study the Galactic gravitational potential: the
        stellar components as well as the dark matter \citep{Johnston2016}.

        On the observational front there has been a large concerted effort in
        the field to find new stellar streams \citep[for an extensive list of
        known Galactic streams see][]{Mateu2022}, extend the detected extents of
        known streams \citep[e.g. for \stream{Palomar 5} alone:][]{Rockosi+2002,
        Ibata+2017, Grillmair+2006, Carlberg+2012, Starkman+2019}, and increase
        the general purity of stream catalogs.

@@

        \begin{enumerate}
            \item Streamfinder \citep{STREAMFINDER}. This method requires the potential
            \item VIA MACHINAE \citep{Shih+2022}
            \item Kiyan and Adrian
            \item Matched Filter
            \item Gaia stuff: Adrian and Ana and Sara et al
            \item Uniform Modeling of 12 Stellar stream \citep{Patrick+2022}. Only on phi1, phi2, and CMD. Their CMD method is more sophisticated. Worth discussing.
        \end{enumerate}

@@

        However, data observations are not taken from the GCOM, but from  $\sim
        8$ kpc distant \citep{GRAVITY2018, Leung+2022} in a local frame,
        typically from vantage points such as Earth or its solar Lagrange
        points, and subsequently represented in the International Celestial
        Reference System (ICRS) \citep{ICRS1997}.  Restricted as we are to a
        very local volume, there is no means to observe systems from the more
        ``intrinsic" reference frames. In our observational frames there is no
        inherent assurance of effecting a rotation that would similarly align
        the stream within observer coordinates such that $\phi_2(\phi_1) \approx
        0$.  Thankfully, most streams are adequately distant from both the
        Galactic center and the observer, or exhibit orientations such that a
        rotation a) can be established or b) holds valid. For most of the
        best-studied globular cluster streams -- \stream{GD-1}, \stream{Palomar
        5}, \stream{Jhelum}, etc. -- this holds true. However for some streams
        of dwarf galaxies, like the Magellanic stream which wraps multiple times
        around the Milky Way \citep{WannierWrixon1972}, no such transformation is
        possible. More generally, a stream's ``sky"-projected path may be
        kinked, e.g. by subhalo interactions, such that it is a many-to-one
        function in $\phi_1$.

        In this work we treat with the vast majority of streams, which have a
        ``sky" transformation such that the stream is an injective function in
        $\phi_2(\phi_1)$.  Since we are examining known streams this
        transformation is known \textit{a priori}, or may be constructed during
        pre-processing \citep[see][\S2.1]{Starkman+2023}.  We note that, with
        sufficient prior knowledge, many-to-one functions may be broken into
        injective segments and their connection constrained by a prior. Thus the
        methods developed subsequently may extented and applied even to wrapped
        and kinked streams. However, we leave extending the model framework to a
        future work.

@@

        Mixture models (MMs) are a statistical method to represent a population
        as a set of a sub-populations specified by probability distributions.
        These distributions are characterized by their parameters and the
        mixture by a corresponding set of mixture coefficients. Scalar mixture
        models may be extended to model completely general conditional
        distributions by making the distribution parameters and mixture
        coefficients into functions of a conditional input
        \citep{McLachlanBasford1989}.  Mixture Density Networks (MDNs) are a
        machine learning method to implement these general MM, taking the
        parameters and coefficients to be the outputs of feed-forward neural
        networks and as such, general continuous functions \citep{Bishop1994}.
        Functional parameters allow the model to capture variation over those
        parameters that is not possible with scalar-valued models. Moreover, by
        using neural networks the inferred functional forms are ``model-free"
        and driven entirely by the data.

@@

        Streams have a diversity of measurement coverage: several have full 6-D
        astrometrics as well as photometrics and chemical abundances available
        \citep[e.g.,][]{Koposov+2019, Antoja+2020, Li+2022},
        while most have partial coverage, lacking some feature dimensions.
        % TODO: terminology ND-phase-space vs feature columns
        Most streams in \Gaia {\color{red} [cite]}, for instance, have proper
        motions but not radial velocities nor reliable parallax measurements.
        Followup surveys, like \dataarchive{S5} {\color{red} [cite]}, add many
        feature dimensions, however only for a select set of streams. Given the
        diversity of coverage it is important that the likelihood model --
        \autoref{eq:membership_prob} -- account for missing data.

@@

            % link to the code!
            In the code we offer a few ansatz for the PDMF, though any
            user-defined function may be used instead. The default assumption,
            known to be non-physical, is a uniform distribution over $\gamma$.
            Similarly non-physical are truncated uniform distributions,
            considering only a segment of the isochrone, e.g. the main sequence
            turnoff (MSTO). We include also a Kroupa initial mass function (IMF)
            \citep{Kroupa2001}. The Kroupa IMF, and all
            physically-motivated IMFs, have an much larger fraction of low-mass
            stars then high-mass stars, the latter being less likely to form.
            This is observed in streams, particularly their progenitors. However
            the Kroupa IMF, and all IMF, are poor fits to the observed PDMF as
            stream progenitors like globular clusters are extremely old
            ($\sim10$ Gyr) while an IMF is the zero-age distribution.  How then
            should the PDMF $\prior_{\cal{I}}$ be treated?
            % ⬆ rephrase previous sentence
            One approach to the PDMF is to use a more informed distribution, for
            example time-evolving an IMF, either analytically or by simulation,
            to the age of the stream's stellar population.  As discussed
            previously, globular clusters are known to be poorly fit by
            theoretic isochrones with realistic inputs (e.g. age, $Z$). Thus
            fitting an isochrone's track to the data means its age, metallicity
            and therefore PDMF are not well fit \TODO{[cite]}. Moreover, for
            many streams, e.g. those lacking known progenitors, the PDMF is
            observationally poorly constrained.  In short, using any fixed
            distribution is simple in implementation, however the choice of
            distribution is not.

@@

        We generate synthetic stream observations by sampling uniformly in $\phi_1$
        In $\phi_2$ we sample from a quadratic in $\phi_1$ with normally-distributed
        $0.15^{\degree}$ scatter.  The distances are sampled from a linear funciton in
        $\phi_1$ with normally-distributed $0.25 \ \unit{kpc}$ scatter, then
        transformed to the parallax $\parallax$.  Non-poissonian density variations
        are then simulated by evaluating the stream density with two Gaussian
        distributions centered at different values of $\phi_1$ with random
        amplitudes (which sum to $1$) and widths.  Stream stars are then removed
        using random sampling, weighted by their likelihood of belonging to the
        density. Post removal, the number of stream stars is
        $\variable{output/mock/nstream_variable.txt}$.
        $\variable{output/mock/nbackground_variable.txt}$ background astrometric points
        are generated from exponential distributions at each $\phi_1$.

        The synthetic stream's photometry is also simulated using a
        \variable{output/mock/isochrone_age_variable.txt},
        [Fe/H] = \variable{output/mock/isochrone_feh_variable.txt} MIST isochrone
        \citep{brutus, Dotter2016, Choi+2016}. The isochrone is truncated just
        short of the giant branch, and is sampled from uniformly along the isochrone
        (i.e., the initial mass function, $P(\gamma)$, is uniform) as well as with
        an intrinsic uniform 0.2~dex scatter orthogonal to the isochrone track. The
        isochrone is shifted by the stream's parallax modulus, which is derived from
        the astrometric distance track.  The background photometric coordinates are
        generated as a 2-dimensional Gaussian with non-zero covariance.  Thus, the
        synthetic data scenario described here generates a mock stream in
        astrometric coordinates and photometric coordinates, with density
        variations, a distance gradient which is consistent in parallax and distance
        modulus (i.e., magnitudes), all superimposed over a noisy field of
        background points in each astrometric and photometric dimension.

@@

        We select on-stream and off-stream regions using the known $\phi_1, \phi_2$
        range of the mock stream.  The upper panel of
        \autoref{fig:mock_data_photometric_background_selection} shows these
        selections in blue and black, respectively.  The lower panel shows the on and
        off-stream selections in photometric coordinates.  In this space it is
        apparent that the on-stream region has two populations, drawn from the
        isochrone and from background points included in the simple $\phi_1, \phi_2$
        box.  In contrast the off-stream region, has only the the background population.
        We use the off-stream population to train a conditional normalizing flow
        (see \autoref{sub:method:photometric_model:off_stream}) on the magnitudes
        $g$ and $r$ and conditioned on $\phi_1$.  The normalizing flow is now a
        non-parametric distribution characterizing the photometric background,
        including its variation over $\phi_1$.  For this mock stream the flexibility
        of the normalizing flow greatly exceeds necessity: the background is
        drawn from a $\phi_1$-invariant multivariate normal, essentially the base
        distribution of the normalizing flow.  It is therefore no surprise that
        the trained normalizing flow provides an excellent characterization of
        the background.  We will later see that the flow performs similarly well
        on complex distributions.  Having trained the background photometric model
        we proceed to build and train the full model.

        Per \autoref{eq:stream_and_bkg_prob} the stream is an independent
        addition of both astrometric and photometric models.
        In astrometrics the stream is a Gaussian track in $\phi_2$ and
        parallax $\varpi$, using the model defined in
        \autoref{ssub:method:astrometric_model:on_stream}. We use a 4-layer 
        \texttt{LinTanh} with shape (1 input, 64 hidden, 4 output), where the 4 outputs
        are $[\mu_{\phi_2}, \sigma_{\phi_2}, \mu_{\varpi}, \sigma_{\varpi}](\phi_1)$.
        The network is small, but sufficiently large to permit 15\% dropout and retain
        robust predictions.
        The photometric model is a distance modulus-shifted isochrone, using the model
        outlined in  \autoref{ssub:method:photometric_model:on_stream}. We use the
        same isochrone parameters as were used to simulate the stream. For this mock
        stream we demonstrate simultaneous astrometric and photometric modeling, linking
        the distance modulus and parallax track, as explained in
        \autoref{ssub:method:linking_to_astrometrics}. Thus the photometric model
        does not have its own network, using instead the outputs
        $[\mu_{\varpi}, \sigma_{\varpi}](\phi_1)$.

        The total background model has two pieces, an astrometric and a photometric model.
        As a pre-training phase the photometric normalizing flow model was fitted to
        an off-stream selection and is now held fixed. The astrometric model
        is split in two components, the first is a uniform distribution
        in $\phi_2$, and the second is an exponential distribution for $\varpi$.
        The uniform distribution does not have any parameters, and thus no neural network.
        The exponential distribution has only the slope parameter and we use a 3-layer 
        \texttt{LinTanh} with shape (1, 32, 1) and 15\% dropout. Background and
        stream are combined in a mixture model whose mixture parameters are a 4-layer 
        (1, 64, 1) \texttt{LinTanh} network. We train the mixture model with an
        \texttt{AdamW} optimizer with learning rate of $5\times10^{-3}$ and a scheduler
        to periodically and temporarily increase the learning rate, encouraging the optimizer
        towards better minima.

@@

        We utilize data from \Gaia{} DR3 for the astrometric coordinates of
        each star. 
        To obtain more accurate color information, we cross-match the \Gaia{}
        field on \PanStarrs{} \citep{PanSTARRS1} using 
        \Gaia's provided \texttt{Best-neighbors} catalog.
        To limit the field to a region known to contain GD-1 we adopt broad cuts
        on the cross-matched data. We perform these rough cuts by requiring
        the data lie within the phase-space cube:
        \begin{itemize}
            \item $\phi_1 \in [-100, 40] \ \rm{\degree}$,
            \item $\phi_2 \in [-9, 5] \rm{\degree}$,
            \item $\varpi \in [-10, 1] \ \rm{mas}$,
            \item $\rm{bp-rp} \in [-1, 3] \ \rm{mag}$
            \item $g, r \in [0, 50] \ \rm{mag}$, and not \texttt{null}.
        \end{itemize}
        where, we use Gala's \citep{gala, galav1.3} \texttt{GD1Koposov10} frame
        \citep{Koposov+2010} to define $\phi_1$, $\phi_2$, and associated proper motions.
        For downloading purposes, we split the query into 10 sub-regions,
        in even $\phi_1$ increments. A plot of the $\phi_1, \phi_2$ regions is included
        in the \href{\GitHubURL{}}{\package{ShowYourWork}} repository.
        Stream-specific coordinates ($\phi_1, \phi_2$) are not included in \Gaia.
        One could embed the coordinate transform in the ADQL \citep{ADQL2.0},
        or define great-circle arcs in ICRS \citep{ICRS1997}. Since only rough
        cuts are required and since the boxes are sufficiently small for a reasonable flat-sky
        approximation  we simply translate the ($\phi_1, \phi_2$) sub-region boxes to ICRS.
        We correct the \PanStarrs{} photometry for dust extinction
        using the \textit{Bayestar19} \citep{Green+2019} models in the \package{dustmaps} package \citep{Green2018}.
        \textit{Bayestar19 is a three-dimensional dust map of the Milky Way
        for which we use \Gaia's parallax distance information.
        We include the $1-\sigma$ errors as a source of uncertainty
        in $\Sigma_n$, the per-star photometric uncertainty.
        % TODO: actually implement that in the model
        The result is a field populated by \variable{output/gd1/ndata_variable.txt} stars, from which we would
        like to identify stream members.

        In \autoref{fig:gd1-data_selection} we create data masks for a more refined
        selection around the stream.

        We make proper motion selection boxes at box corners $(\mu_{\phi_1}, \mu_{\phi_2}) = [(-15,-10) (-4.5, -2)] \ \unit{mas/yr}$, and $(\mu_{\phi_1}, \mu_{\phi_2}) = [(-15,-5) (-5, -1)] \ \unit{mas/yr}$,

        \begin{figure*}[h]
            \script{gd1/plot/data_selection.py}
            \centering
            \includegraphics[width=1\linewidth]{figures/gd1/data_selection.pdf}
            \caption{%
                Data selections for \stream{GD-1}, within the data cube described
                in \autoref{sub:results_gd1:data}.
                %
                \textbf{Row 1 Panels:}
                Proper motion selection boxes in red and yellow.
                The right plot
                %
                \textbf{Row 2 Panels:}
                Photometric selection (blue curve). The selection is based on
                a 12 Gyr, [Fe/H] = -1.3 MIST isochrone, buffered by 0.2 dex.
                % 
                \textbf{Row 3 Panels:}
                Applying the red astrometric and blue photometric selections, plotted
                for astrometric coordinates $\phi_2, \varpi, \mu_{\phi_1}, \mu_{\phi_2}$.
                % 
                \textbf{Row 4 Panels:}
                Applying the yellow astrometric and blue photometric selections, plotted
                for the same astrometric coordinates as the above row.
            }
            \label{fig:gd1-data_selection}
        \end{figure*}

@@

        \todo[inline]{rephrase, this is quoting from the mock stream}
        We select on-stream and off-stream regions using the known $\phi_1,\phi_2$ range
        of the stream. The upper panel of \autoref{fig:gd1-photometric_background_selection}
        shows these selections in blue and black, respectively.
        The lower panel shows the on and off-stream selections in photometric
        coordinates. An isochrone is apparent in the distribution.
        We use the off-stream population to train a
        conditional normalizing flow (see \autoref{sub:method:photometric_model:off_stream})
        on the magnitudes $g$ and $r$ and conditioned on $\phi_1$.
        The normalizing flow is now a non-parametric distribution characterizing
        the photometric background, including its variation over $\phi_1$.

        We model GD-1 as a three-component mixture consisting of a main stream
        component, a spur, and background.

        \todo[inline]{rephrase }
        Per \autoref{eq:stream_and_bkg_prob} the stream is an independent addition
        of both astrometric and photometric models.
        In astrometrics the stream is a Gaussian track in $\phi_2, \mu_{\phi_1}, \mu_{\phi_2}$
        using the model defined in \autoref{ssub:method:astrometric_model:on_stream}.
        We use a 4-layer \texttt{LinTanh} with shape (1, 128, 6), where the 6 outputs
        are $[\mu_{\phi_2}, \sigma_{\phi_2}, \mu_{\mu_{\phi_1}^*}, \sigma_{\mu_{\phi_1}^*}, \mu_{\mu_{\phi_2}}, \sigma_{\mu_{\phi_2}}](\phi_1)$. The network
        is small, but sufficiently large to permit 15\% dropout and retain robust
        predictions. The photometric model is a distance modulus-shifted isochrone,
        using the model outlined in \autoref{ssub:method:photometric_model:on_stream}.
        We use the same isochrone parameters used in the photometric selection.
        The photometric model has its own network.

        In order to guide our model towards
        GD-1, we set down 4 ``control points" along GD-1's main track, and 3
        control points along the spur (illustrated in \TODO{x}). These control
        points act as Gaussian priors on the model. We set the Gaussian error on
        these control points to large values, so that their influence on the
        model is not dominant. Indeed, we find that the value of the control
        points is to help reduce training time, since otherwise it will take the
        flexible density model longer time to converge to the actual stream.

        We also add 3 control points in the distance modulus.
        We draw the points from \TODO{[cite]} and add significant width.

        The background model is an exponential distribution.

        The weight parameter is also set to truncate to $0$ at $\phi_1 < X$
        and $\phi_1 > Y$. This ensures that we do not model stream stars that fall.

@@

@@

        \subsubsection{VIA MACHINAE}{\cite{Shih+2022}}
